Stability concerns in random matrices

### 2018-03-17 random_matrix_stability.Rmd
Rearrange and clean up abstract

**Analyses of dynamic systems using random matrices find that stability generally decreases with increasing system size. But in ecological communities, the rate of change of species density and the magnitude of interaction effects depend on species generation time. Per capita effects of individuals from species with short generation times will be lower than individuals from species with long generation times, and densities of species with shorter generation times will change more rapidly. Interaction effects are therefore expected to be correlated within species, with some species having consistently weaker or stronger effects on others per capita. Here we show that by accounting for variation in generation time among species, community stability is substantially increased. Further, the influence of variation in generation time becomes exponentially more important as community size increases. We suggest that variation in species generation time might be a key contributor to stability of large ecological communities.**

@@

Hence, the goal here is to demonstrate the inherent destabilising effect of rapid evolution, which can also be interpreted as plasticity [@Patel2018], on ecological communities.

The destabilising effect of rapid evolution
--------------------------------------------------------------------------------

@@

Combining the two (work in progress)
--------------------------------------------------------------------------------

```{r, echo = FALSE}
dat       <- read.table(file = "rand_sims_var_gen.txt", header = TRUE);
species   <- dat[,1];
pr_stable <- dat[,2] / (dat[,2] + dat[,4]);
par(mar = c(5, 5, 2, 1));
plot(x = species, y = pr_stable, type = "b", lwd = 2, ylim = c(0, 1),
     xlab = "Species number", ylab = "Proportion stable (n)",
     cex.axis = 1.5, cex.lab = 1.5);
```

Code to print the above is below.

```{r, eval = FALSE, echo = TRUE}
random_comm <- function(dm = 2, iter = 10000){
  stabilised_fastevo <- NULL;
  unstabled_fastevo  <- NULL;
  nn           <- dm * dm;
  tot_stabled  <- 0;
  tot_ustabled <- 0;
  tot_eco_stab <- 0;
  stabilised_fastevo <- NULL;
  unstabled_fastevo  <- NULL;
  while(iter > 0){
    epsil   <- runif(n = dm, min = 1, max = 1000);
    eps_dat <- rep(x = epsil, times = dm);
    eps_mat <- matrix(data = c(epsil, epsil), nrow = dm, ncol = dm, byrow = FALSE);
    stabilised   <- 0;
    unstabilised <- 0;
    Adat <- runif(n = nn, min = -4, max = 4);
    Bdat <- runif(n = nn, min = -4, max = 4);
    Cdat <- runif(n = nn, min = -4, max = 4);
    A    <- matrix(data = Adat, nrow = dm, ncol = dm) * eps_mat;
    B    <- matrix(data = Bdat, nrow = dm, ncol = dm) * eps_mat;
    C    <- matrix(data = Cdat, nrow = dm, ncol = dm) * eps_mat;
    egD  <- 1;
    while(egD >= 0){
      Ddat <- runif(n = nn, min = -4, max = 4);
      D    <- matrix(data = Ddat, nrow = dm, ncol = dm) * eps_mat;
      egD  <- max(Re(eigen(D)$values))
    }
    eco_stable       <- max(Re(eigen(A)$values)) < 0;
    slow_evo_maxE    <- max(Re(eigen(A + B %*% solve(D) %*% (-C))$values));
    stable_fast_evo  <- slow_evo_maxE < 0;
    if(eco_stable == FALSE & stable_fast_evo == TRUE){
      list_ele <- length(stabilised_fastevo$A) + 1;
      stabilised_fastevo$A[[list_ele]] <- A;
      stabilised_fastevo$B[[list_ele]] <- B;
      stabilised_fastevo$C[[list_ele]] <- C;
      stabilised_fastevo$D[[list_ele]] <- D;
      stabilised   <- 1;
    }
    if(eco_stable == TRUE & stable_fast_evo == FALSE){
      list_ele <- length(unstabled_fastevo$A) + 1;
      unstabled_fastevo$A[[list_ele]] <- A;
      unstabled_fastevo$B[[list_ele]] <- B;
      unstabled_fastevo$C[[list_ele]] <- C;
      unstabled_fastevo$D[[list_ele]] <- D;
      unstabilised   <- 1;
    }
    if(stabilised == 1){
      tot_stabled <- tot_stabled + 1;
    }
    if(unstabilised == 1){
      tot_ustabled <- tot_ustabled + 1;
    }
    if(eco_stable == TRUE){
      tot_eco_stab <- tot_eco_stab + 1;
    }
    iter <- iter - 1;
    if(iter %% 1000000 == 0){
        print(c(dm, iter));
    }
  }
  results <- list(ecologically_stable = tot_eco_stab, 
                  stabilised = tot_stabled, destabilised = tot_ustabled,
                  stabilised_all   = stabilised_fastevo, 
                  destabilised_all = unstabled_fastevo);
  return(results);
}

NN       <- 20;
res_tabl <- matrix(data = 0, nrow = (NN-1), ncol = 5);
its      <- 100000000;
for(i in 2:NN){
  res              <- random_comm(dm = i, iter = its); 
  res_tabl[i-1, 1] <- i;                               # Species and traits
  res_tabl[i-1, 2] <- res$ecologically_stable;         # Total stable
  res_tabl[i-1, 3] <- res$destabilised;                # Total destabilised
  res_tabl[i-1, 4] <- its - res_tabl[i-1, 2];          # Total unstable
  res_tabl[i-1, 5] <- res$stabilised;                  # Total stabilised
  print(i);
}
```

>>>>

**Analyses of dynamic systems using random matrices find that stability generally decreases with increasing system size. But in ecological communities, the rate that species densities change as a consequence of species interactions will depend on relative species generation times. Densities of species with relatively short generation times will change more rapidly than densities of species with longer generation times. Interaction effects are therefore expected to be correlated within species, with some species densities changing more rapidly than others in a community. Here we show that by accounting for variation in generation time among species, community stability is substantially increased. The influence of variation in generation time becomes exponentially more important as community size increases. And communities are further stabilised when the interaction effects of long-lived species are especially strong. We suggest that variation in species generation time might be a key contributor to stability of large ecological communities.**

@@

**Hence, the goal here is to demonstrate the inherent destabilising effect of rapid evolution, which can also be interpreted as plasticity [@Patel2018], on ecological communities.**

@@
